---
title: "Predicting Sleep Efficiency: Model Comparison and Explanation"
format: html
execute: 
  echo: false
  warning: false
  message: false
---

## 📁 Comparing Features

```{r}
#| label: load-libraries
#| include: false

library(tidyverse)
library(tidymodels)
library(parsnip)
library(broom)
library(marginaleffects)
library(knitr)
```
```{r}
#| label: load-data
#| cache: true

df <- read_csv("Sleep_Efficiency.csv") |> 
  mutate(
    smoking_status = factor(smoking_status, levels = c("No", "Yes")),
    gender = factor(gender)
  ) |>
  drop_na(sleep_efficiency, age, smoking_status, caffeine_consumption)

```
```{r}
# Fit models
model1 <- linear_reg() |> fit(sleep_efficiency ~ age + smoking_status, data = df)
model2 <- linear_reg() |> fit(sleep_efficiency ~ caffeine_consumption + smoking_status, data = df)
model3 <- linear_reg() |> fit(sleep_efficiency ~ age + caffeine_consumption, data = df)

# Compare models using R-squared
results <- bind_rows(
  glance(model1) |> mutate(model = "age + smoking_status"),
  glance(model2) |> mutate(model = "caffeine + smoking"),
  glance(model3) |> mutate(model = "age + caffeine")
)

# Format as percentages
results <- results |>
  mutate(across(contains("r.squared"), scales::percent_format(accuracy = 0.1)))

# Display table
results |>
  select(model, r.squared, adj.r.squared) |>
  kable(caption = "Model Comparison Based on R-squared (%)")

```

## 📊 Statistical Model

We use a **linear regression model** to estimate how **age** and **smoking status** affect sleep efficiency:

$$
\hat{Y} = \beta_0 + \beta_1 \cdot \text{Age} + \beta_2 \cdot \text{SmokingStatus}_{\text{Yes}} + \epsilon
$$

This model assumes:

- **Y** = sleep efficiency  
- **Age** = age of the participant  
- **SmokingStatus<sub>Yes</sub>** = 1 if smoker, 0 if non-smoker  
- **β₀** = intercept (baseline sleep efficiency)  
- **β₁, β₂** = effect of each predictor  
- **ε** = random error not explained by predictors


```{r}
# Show model coefficients with confidence intervals
tidy(model1, conf.int = TRUE) |>
  select(term, estimate, conf.low, conf.high) |>
  kable(
    caption = "Estimates and 95% Confidence Intervals for Parameters  
    Source: Sleep Efficiency Dataset"
  )

```

```{r}
plot_predictions(model1, condition = "smoking_status", draw = FALSE) |>
  ggplot(aes(x = smoking_status, y = estimate, color = smoking_status)) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high),
                  size = 1.2, linewidth = 1.5, fatten = 3) +
  scale_color_manual(values = c("No" = "#2ECC71", "Yes" = "#E74C3C")) +
  labs(
    title = "Predicted Sleep Efficiency by Smoking Status",
    subtitle = "Smokers tend to have lower predicted sleep efficiency than non-smokers",
    caption = "Source: Sleep Efficiency Dataset",
    x = "Smoking Status",
    y = "Predicted Sleep Efficiency",
    color = "Smoking"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```


<br>
<br>
## 📝 Summary of Analysis

- We compared **three linear models** with different predictors to explain sleep efficiency.
- The **best-performing model** included **age** and **smoking status** as predictors.
- **Smoking** was associated with **lower sleep efficiency**, even after adjusting for age.
- **Age** also had a **small negative effect** on sleep quality.
- These results suggest that both **lifestyle factors (like smoking)** and **demographics (like age)** influence sleep health.
